### 2014-10-09 doc/thesis/system.tex
Adding first versions of the typing rules for statements
Starting to write about subtyping

In the previous chapter we presented an informal overview of Typed Lua.
We showed that programmers can use Typed Lua to combine static and dynamic
typing in the same code.

In this chapter we present the formalization of Typed Lua.
We use typing rules to show how we implemented the features that we
discussed in the previous chapter.

@@

@@

@@

\begin{figure}[!ht]
\textbf{Abstract Syntax}\\
\dstart
$$
\begin{array}{rl}
s ::= & \;\; \mathbf{skip} \; | \;
s_{1} \; ; \; s_{2} \; | \;
\vec{l} = \vec{e}  \; | \;
\mathbf{while} \; e \; \mathbf{do} \; s \; | \;
\mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}\\
& | \; \mathbf{local} \; \vec{n}{:}\vec{t} = \vec{e} \; \mathbf{in} \; s \; | \;
\mathbf{rec} \; n{:}t = f \; \mathbf{in} \; s \; | \;
\mathbf{return} \; \vec{e} \; | \;
a\\
e ::= & \;\; \mathbf{nil} \; | \;
\mathbf{false} \; | \;
\mathbf{true} \; | \;
{\it number} \; | \;
{\it string} \; | \;
{...} \; | \;
f \; | \;
\{ \; \vec{c} \; \} \\
& | \; e_{1} + e_{2} \; | \;
e_{1} \; {..} \; e_{2} \; | \;
e_{1} == e_{2} \; | \;
e_{1} < e_{2} \; | \;
e_{1} \; \mathbf{and} \; e_{2} \; | \;
e_{1} \; \mathbf{or} \; e_{2} \\
& | \; \mathbf{not} \; e \; | \;
- e \; | \;
\# \; e \; | \;
a \; | \;
l \; | \;
(e) \; | \;
{<}t{>} \; e\\
l ::= & \; n \; | \;
e_{1}[e_{2}]\\
a ::= & \; e_{1}(\vec{e_{2}}) \; | \;
e_{1}{:}n(\vec{e_{2}})\\
f ::= & \; \mathbf{fun} \; (){:}r \; s \; | \;
\mathbf{fun} \; ({...}{:}t){:}r \; s \; | \;
\mathbf{fun} \; (\vec{n}{:}\vec{t}){:}r \; s \; | \;
\mathbf{fun} \; (\vec{n}{:}\vec{t},{...}{:}t){:}r \; s\\
c ::= & \; [e_{1}] = e_{2}\\
n ::= & \; {\it name}\\
\end{array}
$$
\dend
\caption{Typed Lua abstract syntax}
\label{fig:syntax}
\end{figure}

>>>>

In the previous chapter we presented an informal overview of Typed Lua.
We showed that programmers can use Typed Lua to combine static and dynamic
typing in the same code, allowing them to incrementally migrate from
dynamic to static typing.
This is a benefit to programmers that use dynamically typed languages
to build large applications, as static types detect many bugs
during the development phase, and also provide better documentation.

In this chapter we present the formalization of Typed Lua.
Besides its practical contribution, Typed Lua also has some interesting
contributions on the field of optional type systems for scripting
languages that we explain in this chapter.
Namely, it is a simple type system that covers several Lua idioms
and features, such as the refinement of tables, multiple return values,
and optional parameters.
We use typing rules to explain how we implemented the features of Typed Lua.

@@

Our type system uses subtyping \citep{abadi1996to,cardelli1984smi} to order
types and consistent-subtyping \citep{siek2007objects,siek2013mutable}
to allow the interaction between statically and dynamically typed code.
We will explain these two relations along this section.
Firt, we will explain the subtyping relationship while we introduce
each type that is present on our type system.
Then, we will explain the differences between subtyping and
consistent-subtyping, showing that they are orthogonal relations.

@@

Literal types represent the type of boolean, number, and string
constants.
We shall see that literal types are important to type tables as
records.
The following subtyping rule shows that each literal type
is a subtype of itself:
\[
\begin{array}{c}
\mylabel{S-LITERAL}\\
\senv \vdash l \subtype l
\end{array}
\]

The variable $\senv$ is a set of pairs of recursion variables that
records the pairs of valid recursive types.

The base types $\Boolean$, $\Number$, and $\String$ represent the
type of values that during run-time Lua tags as booleans, numbers,
and strings, respectively.
Literal types are subtypes of their respective base types, and
each base type is a subtype of itself, as the following subtyping
rules show:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FALSE}\\
\senv \vdash \False \subtype \Boolean
\end{array}
\;
\begin{array}{c}
\mylabel{S-TRUE}\\
\senv \vdash \True \subtype \Boolean
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-NUMBER}\\
\senv \vdash {\it number} \subtype \Number
\end{array}
\;
\begin{array}{c}
\mylabel{S-STRING}\\
\senv \vdash {\it string} \subtype \String
\end{array}
\;
\begin{array}{c}
\mylabel{S-BASE}\\
\senv \vdash b \subtype b
\end{array}
\end{array}
\]

The type $\Nil$ represents both the type of the constant \texttt{nil}
and the type of values that during run-time Lua tags as nil.
It is a separated type because in our type system $\Nil$ is
not the bottom type.
Typed Lua has this behavior because in Lua the tag \texttt{nil}
represents the absence of an useful value.
Thus, making $\Nil$ the bottom type would severally reduce the
amount of type safety in our type system.
The next subtyping rule shows this behavior:
\[
\begin{array}{c}
\mylabel{S-NIL}\\
\senv \vdash \Nil \subtype \Nil
\end{array}
\]

The type $\Value$ is the top type, so any first-level type is a
subtype of $\Value$.
We shall see that this type along with variadic types help to
preserve the semantics of Lua to drop extra values on multiple
assignments, and function calls.
Following we show its subtyping rule:
\[
\begin{array}{c}
\mylabel{S-VALUE}\\
\senv \vdash t \subtype \Value
\end{array}
\]

The dynamic type $\Any$ is the type that allows programmers to mix
static and dynamic typing.
The type $\Any$ is neither the bottom nor the top type, but a
separated typed that is a subtype only of itself:
\[
\begin{array}{c}
\mylabel{S-ANY}\\
\senv \vdash \Any \subtype \Any
\end{array}
\]

If we set the type $\Any$ as both bottom and top types of our type
system, then all programs would type check without errors.
This would happen due to the transitivity of subtyping, that is,
we would be able to down cast any type to $\Any$ and then up cast
$\Any$ to any type.
We shall see later in this section that we use consistent-subtyping
to allow the dynamic type to interact with other first-level types,
and thus allow dynamically typed code coexist with statically typed
code.

Typed Lua uses the type $\Self$ to represent the type of the \texttt{self}
parameter in object-oriented Lua code.
It is only a subtype of itself:
\[
\begin{array}{c}
\mylabel{S-SELF}\\
\senv \vdash \Self \subtype \Self
\end{array}
\]

Union types represent types that can hold a value of several
different, but fixed types.
Notice that union types are disjoint, that is, all types in
the union are not subtypes or supertypes among themselfs.
Next we show the subtyping rules for union types:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-UNION1}\\
\dfrac{\senv \vdash t_{1} \subtype t \;\;\;
       \senv \vdash t_{2} \subtype t}
      {\senv \vdash t_{1} \cup t_{2} \subtype t}
\end{array}
\;
\begin{array}{c}
\mylabel{S-UNION2}\\
\dfrac{\senv \vdash t \subtype t_{1}}
      {\senv \vdash t \subtype t_{1} \cup t_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-UNION3}\\
\dfrac{\senv \vdash t \subtype t_{2}}
      {\senv \vdash t \subtype t_{1} \cup t_{2}}
\end{array}
\end{array}
\]

The first rule shows that a union type $t_{1} \cup t_{2}$
is subtype of $t$ if both $t_{1}$ and $t_{2}$ are subtypes
of $t$;
while the other rules show that a type $t$ is subtype
of a union type $t_{1} \cup t_{2}$ if $t$ is subtype of
$t_{1}$ or if $t$ is subtype of $t_{2}$.

To introduce function types we first need to introduce
second-level types, as they appear on both sides of function types.
Basically, second-level types are tuples of first-level types
that can optionally end with a variadic type.
Our type system needs second-level types because tuples are
not first-class values in Lua, though they may appear on
argument passing, multiple returns, and multiple assignments.
A variadic type $t*$ is a generator for a sequence of values of
type $t \cup \Nil$, and it represents the type of the vararg
expression.
Next we show their subtyping rules:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-VARARG}\\
\dfrac{\senv \vdash t_{1} \cup \Nil \subtype t_{2} \cup \Nil}
      {\senv \vdash t_{1}* \subtype t_{2}*}
\end{array}
\;
\begin{array}{c}
\mylabel{S-TUPLE1}\\
\dfrac{\senv \vdash t_{1} \subtype t_{2} \;\;\;
       \senv \vdash s_{1} \subtype s_{2}}
      {\senv \vdash t_{1} \times s_{1} \subtype t_{2} \times s_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-TUPLE2}\\
\dfrac{\senv \vdash t_{1} \cup \Nil \subtype t_{2} \;\;\;
       \senv \vdash t_{1}* \subtype s_{2}}
      {\senv \vdash t_{1}* \subtype t_{2} \times s_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-TUPLE3}\\
\dfrac{\senv \vdash t_{1} \subtype t_{2} \cup \Nil \;\;\;
       \senv \vdash s_{1} \subtype t_{2}*}
      {\senv \vdash t_{1} \times s_{1} \subtype t_{2}*}
\end{array}
\end{array}
\]

The subtyping rules for variadic and tuple types show that
these types are covariant.
The subtyping rule for variadic types is straightforward:
it shows that a variadic type $t_{1}*$ is subtype of another
variadic type $t_{2}*$ if $t_{1} \cup \Nil$ is subtype of
$t_{2} \cup \Nil$.
The first rule for tuple types shows that a tuple type
$t_{1} \times s_{1}$ is subtype of another tuple type
$t_{2} \times s_{2}$ if $t_{1}$ is subtype of $t_{2}$
and $s_{1}$ is subtype of $s_{2}$.
As we mentioned, a tuple can optionally end with a variadic
type, and the second and third rules for tuple types handle
this case.
The second rule shows that a variadic type is subtype of
a tuple whenever the type of the value that it generates is subtype
of all the elements of the tuple.
The third rule shows that a tuple is subtype of a variadic
type whenever all the elements of the tuple are subtype of the
type of the value that it generates.
We shall see that these subtyping relations along with the
types $\Value$ and $\Nil$ allow our type system to reflect
the semantics of Lua on discarding extra parameters and
replacing missing parameters.

@@

\begin{figure}[!ht]
\textbf{Abstract Syntax}\\
\dstart
$$
\begin{array}{rl}
s ::= & \;\; \mathbf{skip} \; | \;
s_{1} \; ; \; s_{2} \; | \;
\vec{l} = \vec{e}  \; | \;
\mathbf{while} \; e \; \mathbf{do} \; s \; | \;
\mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}\\
& | \; \mathbf{local} \; \vec{n{:}t} = \vec{e} \; \mathbf{in} \; s \; | \;
\mathbf{rec} \; n{:}t = f \; \mathbf{in} \; s \; | \;
\mathbf{return} \; \vec{e} \; | \;
a\\
e ::= & \;\; \mathbf{nil} \; | \;
\mathbf{false} \; | \;
\mathbf{true} \; | \;
{\it number} \; | \;
{\it string} \; | \;
{...} \; | \;
f \; | \;
\{ \; \vec{c} \; \} \\
& | \; e_{1} + e_{2} \; | \;
e_{1} \; {..} \; e_{2} \; | \;
e_{1} == e_{2} \; | \;
e_{1} < e_{2} \; | \;
e_{1} \; \mathbf{and} \; e_{2} \; | \;
e_{1} \; \mathbf{or} \; e_{2} \\
& | \; \mathbf{not} \; e \; | \;
- e \; | \;
\# \; e \; | \;
a \; | \;
l \; | \;
(e) \; | \;
{<}t{>} \; e\\
l ::= & \; n \; | \;
e_{1}[e_{2}]\\
a ::= & \; e_{1}(\vec{e_{2}}) \; | \;
e_{1}{:}n(\vec{e_{2}})\\
f ::= & \; \mathbf{fun} \; (){:}r \; s \; | \;
\mathbf{fun} \; ({...}{:}t){:}r \; s \; | \;
\mathbf{fun} \; (\vec{n{:}t}){:}r \; s \; | \;
\mathbf{fun} \; (\vec{n{:}t},{...}{:}t){:}r \; s\\
c ::= & \; [e_{1}] = e_{2}\\
n ::= & \; {\it name}\\
\end{array}
$$
\dend
\caption{Typed Lua abstract syntax}
\label{fig:syntax}
\end{figure}
